---
title: "Wintermoth"
format: html
editor: visual
---

## Code for extracting land surface phenology from VIIRS GLSP for wintermoth sample location

This code is used to extract Land surface phenology (LSP) from VIIRS Global Land Surface Phenology. See <https://lpdaac.usgs.gov/documents/2263/VNP22_UserGuide_v2.pdf> for the VIRRS GLSP user guide.

Data is available for the period 2013-2023

## Libraries

```{r}
library(renv)
renv::init()
```

```{r echo=FALSE}
install.packages("dplyr")
install.packages("terra")
install.packages("ggplot2")
install.packages("ggspatial")
install.packages("hdf5r")

library(dplyr) #For data wrangling 
library(terra) #For raster data 
library(ggplot2) #For data visualization 
library(ggspatial) #For data visualization 
library(hdf5r)

```

## Load data

create list of files available for the downloaded VIIRS GLSP data

```{r echo=FALSE}
# create list of files from the downloaded VIIRS data
files <- list.files("../../data/VIIRS", pattern = "VNP22Q2.*\\.h5$", full.names = TRUE)

# Check the tile names to confirm you have multiple tiles
basename(files)

# Extract year from each filename
years <- as.integer(substr(basename(files), 10, 13)) # extract the name based on the filename
```

Since we are not interested in all the bands within a raster, we need to specify the ones we are interested in.

Onset of greenness band

-   Onset_greenness_increase band, the date of onset if greenness increase (often referred to as greenup onset, start of growing season, or start of spring).

Furthermore we would like to include some of the quality control bands

-   PCQ_Onset_greenness_Increase, PCQ around greenup onset, which is calculates as the proportion of 3-day good observations during the periods of 3 3-day EVI2 composite before and after the greenup onset

-   GLSP_QC which specifies the overall quality of the product and the proportion of climatology used in a growing-season time series.

Create function to load the relevant bands

```{r echo=FALSE}

# function to extract relevant bands
extract_vnp <- function(filepath, year) {
  
  sinu_crs  <- "+proj=sinu +lon_0=0 +x_0=0 +y_0=0 +a=6371007.181 +b=6371007.181 +units=m +no_defs"
  tile_size <- 1111950.5196
  
  # Parse tile h and v from filename dynamically
  tile_str <- regmatches(basename(filepath), regexpr("h\\d{2}v\\d{2}", basename(filepath)))
  tile_h   <- as.integer(substr(tile_str, 2, 3))
  tile_v   <- as.integer(substr(tile_str, 5, 6))
  
  xmin <- (tile_h - 18) * tile_size
  xmax <- xmin + tile_size
  ymax <- (9 - tile_v) * tile_size
  ymin <- ymax - tile_size
  tile_ext <- ext(xmin, xmax, ymin, ymax)
  
  # Build paths to relevant bands within each raster
  onset_path <- paste0('HDF5:"', filepath, '"://HDFEOS/GRIDS/Cycle_1/Data_Fields/Onset_Greenness_Increase_1')
  qc_path    <- paste0('HDF5:"', filepath, '"://HDFEOS/GRIDS/Cycle_1/Data_Fields/GLSP_QC_1')
  pgq_path   <- paste0('HDF5:"', filepath, '"://HDFEOS/GRIDS/Cycle_1/Data_Fields/PGQ_Onset_Greenness_Increase_1')
  
  # Load and flip vertically to correct upside-down orientation
  onset <- flip(rast(onset_path), direction = "vertical")
  qc    <- flip(rast(qc_path),    direction = "vertical")
  pgq   <- flip(rast(pgq_path),   direction = "vertical")
  
  # Assign extent and CRS
  ext(onset) <- tile_ext;  crs(onset) <- sinu_crs
  ext(qc)    <- tile_ext;  crs(qc)    <- sinu_crs
  ext(pgq)   <- tile_ext;  crs(pgq)   <- sinu_crs
  
  # Correct values
  offset <- (year - 2000) * 366 # greenup_onset is given in days after 2000 (need to correct for this which is what we are doing by creating this offset)
  
  onset_corrected <- app(onset, function(x) ifelse(x == 32767, NA, x - offset)) # correct green-up using offset
  qc_clean        <- app(qc,    function(x) ifelse(x == 255,   NA, x))
  pgq_clean       <- app(pgq,   function(x) ifelse(x == 255,   NA, x))
  
  ext(onset_corrected) <- tile_ext;  crs(onset_corrected) <- sinu_crs
  ext(qc_clean)        <- tile_ext;  crs(qc_clean)        <- sinu_crs
  ext(pgq_clean)       <- tile_ext;  crs(pgq_clean)       <- sinu_crs
  
  names(onset_corrected) <- paste0("Onset_Greenness_Increase_", year)
  names(qc_clean)        <- paste0("GLSP_QC_", year)
  names(pgq_clean)       <- paste0("PGQ_Onset_Greenness_Increase_", year)
  
  return(c(onset_corrected, qc_clean, pgq_clean))
}

```

Use function to load in relevant bands for each year and each tile (4 different tiles used for this projects)

```{r echo=FALSE}
# Process all files individually first
all_results <- mapply(extract_vnp, filepath = files, year = years, SIMPLIFY = FALSE)
names(all_results) <- paste0(substr(basename(files), 10, 13), "_", 
                             regmatches(basename(files), regexpr("h\\d{2}v\\d{2}", basename(files))))

# e.g. names will be like "2013_h17v03", "2013_h17v04", giving information on the year as well as the tile 
unique_years <- unique(years)


```

Create a function to combine tiles from the same year for easier data extraction

```{r echo=FALSE}

mosaicked <- lapply(unique_years, function(yr) {
  
  # Get all tiles for this year
  yr_idx  <- which(years == yr)
  yr_tiles <- all_results[yr_idx]
  
  message("Mosaicking ", length(yr_tiles), " tiles for year ", yr)
  
  # Mosaic each layer by creating a SpatRasterCollection first
  onset_mosaic <- mosaic(sprc(lapply(yr_tiles, function(r) r[[1]])))
  qc_mosaic    <- mosaic(sprc(lapply(yr_tiles, function(r) r[[2]])))
  pgq_mosaic   <- mosaic(sprc(lapply(yr_tiles, function(r) r[[3]])))
  
  # Name layers
  names(onset_mosaic) <- paste0("Onset_Greenness_Increase_", yr)
  names(qc_mosaic)    <- paste0("GLSP_QC_", yr)
  names(pgq_mosaic)   <- paste0("PGQ_Onset_Greenness_Increase_", yr)
  
  return(c(onset_mosaic, qc_mosaic, pgq_mosaic))
})

names(mosaicked) <- unique_years
```

#### Wintermoth data

load in wintermoth data and ensure that it is in the same projection as the VIIRS data

```{r}
wm_location <- read.csv("../../data/wgs_europe_metadata.csv")

location_vect <- vect(wm_location, geom = c("longitude", "latitude"), crs = "EPSG:4326")

location_sin <- project(location_vect, "+proj=sinu +lon_0=0 +x_0=0 +y_0=0 +a=6371007.181 +b=6371007.181 +units=m +no_defs")
```

## Extract phenology data

```{r echo=FALSE}

all_extracted <- lapply(names(mosaicked), function(yr) {
  
  extracted <- extract(mosaicked[[yr]], location_sin)
  
  # Rename extracted columns to remove year suffix so they match across years
  names(extracted) <- c("ID", "Onset_Greenness_Increase", "GLSP_QC", "PGQ_Onset_Greenness_Increase")
  
  # Drop the ID column (just a row index added by extract())
  extracted <- extracted[, -1]
  
  out <- cbind(wm_location, year = yr, extracted)
  return(out)
})

final_table <- do.call(rbind, all_extracted)

head(final_table)


```

## Plot one year of data (2013)

```{r, echo=TRUE, fig.width=7, fig.height=6}
r <- mosaicked[["2013"]][[1]]

plot(r,
     main = "Onset of Greenness 2013 - Mosaicked",
     col  = hcl.colors(100, "RdYlGn"))

# safest if location_sin is a spatial object:
plot(location_sin, add = TRUE, col = "green", pch = 16, cex = 0.8)
```
